---
author: "Simon Garnier"
title: "3 - Manipulating a track table"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{manipulate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r paged.print=FALSE}
library(trackdf)
data("tracks")
```

## 3.1 - Creating track tables

Under the hood, a track table is just a simple data frame of class 
[`data.frame`](https://stat.ethz.ch/R-manual/R-patched/library/base/html/data.frame.html), 
[`tibble`](https://CRAN.R-project.org/package=tibble) or 
[`data.table`](https://cran.r-project.org/package=data.table) with a few extra 
bells and whistles. This means that you can manipulate a track table same way 
you would a data frame of any of these three classes. 

To create a track table based on `data.frame`, do: 

```{r paged.print=FALSE}
t_df <- track(x = tracks$x, y = tracks$y, t = tracks$t, id = tracks$id,
  proj = "+proj=longlat", tz = "Africa/Windhoek", table = "df")
head(t_df)
```

To create a track table based on `tibble`, do: 

```{r paged.print=FALSE}
t_tbl <- track(x = tracks$x, y = tracks$y, t = tracks$t, id = tracks$id,
  proj = "+proj=longlat", tz = "Africa/Windhoek", table = "tbl")
t_tbl 
```

To create a track table based on `data.table`, do: 

```{r paged.print=FALSE}
t_dt <- track(x = tracks$x, y = tracks$y, t = tracks$t, id = tracks$id,
  proj = "+proj=longlat", tz = "Africa/Windhoek", table = "dt")
t_dt
```

---

## 3.2 - Basic manipulations

You can check whether an object is a track table as follows:

```{r}
is_track(t_df)
is_track(t_tbl)
is_track(t_dt)
```

You can check whether a track table contains geographic coordinates or not as 
follows:

```{r}
is_geo(t_df)
is_geo(t_tbl)
is_geo(t_dt)
```

---

## 3.3 - Accessing data

You can access the different parts (rows, columns, elements) of a track table as 
follows:

```{r paged.print=FALSE}
head(t_df$id)
head(t_tbl$id)
head(t_dt$id)
```

```{r paged.print=FALSE}
head(t_df[["t"]])
head(t_tbl[["t"]])
head(t_dt[["t"]])
```

```{r paged.print=FALSE}
head(t_df[1])
t_tbl[1]
t_dt[1]
```

```{r paged.print=FALSE}
t_df[1, ]
t_tbl[1, ]
t_dt[1, ]
```

```{r paged.print=FALSE}
t_df[1, 1]
t_tbl[1, 1]
t_dt[1, 1]
```

Note that the results varies depending on the underling data frame class.  

By combining the commands above with the `<-` operator, you can easily modify 
the content of a track table. For example:

```{r}
t_df$id[t_df$id == "1"] <- "0"
head(t_df)
```

```{r}
t_df[t_df[, 1] == "0", 1] <- "1"
head(t_df)
```

---

## 3.4 - Accessing and modifying the coordinates' projection

You can access the projection of a track table as follows. 

```{r}
projection(t_df)
```

You can modify in place the projection of a track table as follows. This will 
automatically convert the coordinates in the appropriate projection system. 

```{r}
projection(t_df) <- "+proj=somerc +lat_0=46.9524056 +lon_0=7.43958333 +ellps=bessel +x_0=2600000 +y_0=1200000 +towgs84=674.374,15.056,405.346 +units=m +k_0=1 +no_defs"
head(t_df)
```

```{r}
projection(t_df) <- "+proj=longlat" 
head(t_df)
```

If you prefer not to modify the original object, you can create a new one with 
the new projection as follows. 

```{r}
t_df_new_proj <- project(t_df, "+proj=somerc +lat_0=46.9524056 +lon_0=7.43958333 +ellps=bessel +x_0=2600000 +y_0=1200000 +towgs84=674.374,15.056,405.346 +units=m +k_0=1 +no_defs")

head(t_df_new_proj)
```

---

## 3.5 - Tidyverse

Track tables are mostly compatible with the functions from the 
[tidyverse](https://www.tidyverse.org/) (with one notable exception, see below).
For instance, you can use all the [`dplyr`](https://dplyr.tidyverse.org/) verbs 
to filter, mutate, group, etc., a track table, the same way you would do with a
`tibble` or a `data.frame`. 

```{r}
if (requireNamespace("dplyr", quietly = TRUE)) {
  library(dplyr)
  
  t_df %>%
    filter(., t < as.POSIXct("2015-09-10 07:01:00", tz = "Africa/Windhoek")) %>%
    head(.)
}
```

```{r}
if (requireNamespace("dplyr", quietly = TRUE)) {
  library(dplyr)
  
  t_df %>%
    group_by(., id) %>%
    summarize(., x = mean(x),
              y = mean(y))
}
```

`bind_rows` is a notable exception to the rule above. For technical reasons in 
the implementation of `bind_rows`, it cannot retain all the attributes of objects
that are not pure `tibble` or `data.frame` (e.g., track tables have an extra 
attribute to store the projection of the coordinates, which is stripped by 
`bind_rows`). For this reason, `trackdf` provides a special version of `bind_rows`
called `rbind_track` to achieve the same purpose. 

With `data.frame` based track tables:

```{r eval=FALSE}
if (requireNamespace("dplyr", quietly = TRUE)) {
  t_df1 <- dplyr::filter(t_df, id == "1")
  t_df2 <- dplyr::filter(t_df, id == "2")
} else {
  t_df1 <- t_df[t_df$id == "1", ]
  t_df2 <- t_df[t_df$id == "2", ]
}

head(rbind_track(t_df1, t_df2))
```

With `tibble` based track tables:

```{r paged.print=FALSE, eval=FALSE}
if (requireNamespace("dplyr", quietly = TRUE)) {
  t_tbl1 <- dplyr::filter(t_tbl, id == "1")
  t_tbl2 <- dplyr::filter(t_tbl, id == "2")
} else {
  t_tbl1 <- t_tbl[t_tbl$id == "1", ]
  t_tbl2 <- t_tbl[t_tbl$id == "2", ]
}

rbind_track(t_tbl1, t_tbl2)
```

With `data.table` based track tables:

```{r paged.print=FALSE, eval=FALSE}
if (requireNamespace("dplyr", quietly = TRUE)) {
  t_dt1 <- dplyr::filter(t_dt, id == "1")
  t_dt2 <- dplyr::filter(t_dt, id == "2")
} else {
  t_dt1 <- t_dt[t_dt$id == "1", ]
  t_dt2 <- t_dt[t_dt$id == "2", ]
}

rbind_track(t_dt1, t_dt2)
```

---

## 3.4 - Plotting

You can use any plotting method accepting a `data.frame`, a `tibble` or a 
`data.table` to represent the data in a track table. 

With base R: 

```{r}
plot(y ~ x, data = t_dt[t_dt$id == "1"], type = "l", col = "red", 
     xlim = range(t_dt$x), lwd = 2, asp = 1)
lines(y ~ x , data = t_dt[t_dt$id == "2"], col = "blue",  lwd = 2)
```

With [`ggplot2`]{https://ggplot2.tidyverse.org/}

```{r}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(data = t_dt, mapping = ggplot2::aes(x = x, y = y, color = id)) +
    ggplot2::geom_path() +
    ggplot2::coord_map()
} 
```
